Fractional Types

### 2019-10-30 fracGC/p.tex
more on intro

\paragraph*{Conservation of Information.} If quantum field theory is
correct (as it so far seems to be) then information, during any
physical process, is neither created nor
destroyed. Landauer~\cite{Landauer:1961,Landauer,bennett1985fundamental},
Bennett~\cite{bennett2010notes,bennett2003notes,Bennett:1973:LRC},
Fredkin~\cite{fredkin1982conservative} and others made compelling
arguments that this physical principle induces a corresponding
computational principle of ``conservation of information.'' This
principle characterizes quantum computing and its classical
restriction known as reversible computing.

\paragraph*{Quantum (Reversible) Computing.} The quantum circuit model
can be characterized as a classical reversible circuit model augmented
with parallel operations modeling superpositions. Computations in the
classical submodel can be formalized as isomorphisms on finite types,
i.e., the types generated from the empty type 0, the unit type 1, and
sums and products. Put differently, the principle of conservation of
information is expressed as computating via sound and complete type
isomorphisms~\cite{James:2012:IE:2103656.2103667} or type
equivalences~\cite{DBLP:conf/esop/CaretteS16}.

>>>>

\paragraph*{Conservation of Information.} If quantum field theory is
correct (as it so far seems to be) then information, during any
physical process, is neither created nor
destroyed. Landauer~\cite{Landauer:1961,Landauer,bennett1985fundamental},
Bennett~\cite{bennett2010notes,bennett2003notes,Bennett:1973:LRC},
Fredkin~\cite{fredkin1982conservative} and others made compelling
arguments that this physical principle induces a corresponding
computational principle of ``conservation of information.'' This
principle is indeed one of the defining characteristics of quantum
computing and its classical restriction known as reversible computing.

\paragraph*{Quantum and Reversible Computing.} The Toffoli gate is
known to be universal for classical reversible
circuits~\cite{Toffoli:1980}. Adding just one gate (the Hadamard gate)
produces a universal set of primitives for quantum
circuits~\cite{hadtoffuniv}. The ``only'' difference between the two
circuit models is that quantum circuits can process superpositions of
values (waves) in one step whereas classical circuits lack this form
of parallelism. Most importantly, structurally the two circuits models
are identical and one can derive properties valid for both families by
focusing on the simpler classical model.

Instead of using the Toffoli gate as a universal primitive for
reversible classical circuits, one can exploit the full force of type
theory and category theory by expressing reversible classical
computations as \emph{isomorphisms over finite
  types}~\cite{Fiore:2004,James:2012:IE:2103656.2103667} or
\emph{equivalences over
  groupoids}~\cite{DBLP:conf/esop/CaretteS16}. This perspective
exposes interesting mathematical structure in reversible computations
that we will exploit to solve the ``ancilla problem'' explained next.
